---
title: "Introduction to R"
author: "Maja Kuzman"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: inverse, center, middle

# R, Rstudio, Rmd

---
### How can we use R and RStudio?  

- console   
- script   
- notebook:   
     new chunk: ctrl + ALT + i 

--
### What can we do with R? 

-  We can use R as a calculator!  

### Try it:  
     
Add a new chunk to line 66.   
Calculate the result of 13%/%3 and 13%%3. 

--

*We can even do better than that...* 

---
class: inverse, center, middle

# Variables and data structures

---
### Variables 


```{r}
myFirstNumber <- 0.1
myFirstVector <- c(2, 3, 7, 8)

```

--

-> Environment  
-> call the variable by name  
-> print() function  
-> one line, part of code shaded and CTRL +ENTER

```{r}
myFirstNumber
print(myFirstNumber)

```

---
class: middle, center

### Data structures in R  

vectors  
list
matrix  
data.frame  
factors  

---
### Vectors 

```{r}
nVec <- c(1, 5, 7, 9, 12.5) # numeric vector
cVec <- c("a", "b", "some words") # character vector
lVec <- c(TRUE, FALSE, T, T, F) # logical vector

vvec <- c(1,5, "word")
vvec
```

```{r}
nVec
cVec
lVec
```

---

### Matrices  

Matrices are tables that have rows and columns and store elements of same type.
```{r}
y <- matrix(1:20, nrow=5,ncol=4)
y
```


---

### Data frames

```{r}
mydf <- data.frame( firstColumn = c(1,5,7),
                    secondColumn = c("a","b","third Element"))
mydf
```

---

### Lists


```{r}
l <- list(firstElement = c(1,5,44,6),
          second = c("a", 3),
          y)
l

```

---

### Factors  

If something is weird, factors are the usual suspects.. 


```{r}
xx <- factor(sample(1:15,20, replace = T))
xx
```

--

If you want a normal numeric vector:
```{r}
as.numeric(xx)

```

--

you should do this... 

```{r}
as.numeric(as.character(xx))

```

---
class: center, middle, inverse  


# Vectors

## The basics

---
### What we can do with vectors:  

#### Make a vector c()  


You can make a vector by using the function c() (concatenate). Here is an example of vectors myFirstvector, and myFirstSequence:

```{r}
myFirstVector <- c("some words","p","word", "last one")
myFirstSequence <- 1:4

```

---

## Subsetting [ ]

Print the whole vector

```{r}
myFirstVector
myFirstSequence

```

Print third element in a vector
```{r}
myFirstVector[3]
myFirstSequence[3]

```

---

### Access multiple elements:  

1) BY POSITION  

Provide a vector of positions to look at:  
```{r}
myFirstVector
```

```{r}
myFirstVector[c(1,3)]


```

```{r}
somePositions <- c(1,3)
somePositions
```

```{r}
myFirstVector[somePositions]

```

---

### Access multiple elements:  

1) BY INCLUSION  

Provide a LOGICAL vector :  
```{r}
myFirstVector
myFirstVector[c(TRUE, FALSE, TRUE,FALSE)]

```

```{r}
someLogicalVector <- c(TRUE,TRUE, FALSE, FALSE)
someLogicalVector
```

```{r}
myFirstVector[someLogicalVector]

```


---

### Exercise! 

1. Create a vector named myvector that contains numbers 15,16,17,18 and 20.  
2. Get 2nd and 4th number in the vector by subsetting.  
3. Get 2nd and 4th number in the vector by providing a logical vector.  


Solution: 
CTRL+Alt+i will insert a new chunk  


---

### Think about it!

#### What happened here??:  

```{r, eval=FALSE}
someothervector <- c(1,0,1,0,1)
myFirstVector[someothervector]
myFirstVector[as.logical(someothervector)]

```

--

```{r, echo=FALSE}
someothervector <- c(1,0,1,0,1)
myFirstVector[someothervector]
myFirstVector[as.logical(someothervector)]

```

---

### Basic operation on vectors  

Same as on numbers:  
\+  
\-  
/  
\*  

Example:   

#### Multiplication  by constant  

```{r}
someothervector * 0.5

```

---

#### Multiplication by other vector:

I) SAME size  

```{r}
someothervector
someothervector * 1:5

```

II) DIFFERENT size : recycling *because why not.*

```{r}
someothervector*c(0.3,0.1)
c(1,2,3,4,5,6) * 1:2 # doesnt produce a warning
```

---

### Basic comparisons  

The following will return a logical vector for every compared position:

```{r}
someothervector == 1
```

```{r}
someothervector == c(1,0,1,0,1)

```

```{r}
someothervector>0

```

--

#### what happened here?  

```{r}
someothervector[someothervector>0]

```

---
### Exercise: Multiplication, recycling and comparison.   
1. Multiply your myvector by c(0.1, 0.2)  
     -> what do you expect to get??
2. Check if you get what you expected by comparing it to vector you expect to get :)   


Solution:  




---

### Exercise: Multiplication, recycling and comparison.   
1. Return only ELEMENTS in myvector that are smaller then 17.
     -> hint - use subsetting and comparison. 


```{r, eval=FALSE}
myvector[ ??? ]
```




---

### Logical operators

AND: & 

```{r, echo=FALSE}
data.frame('first_second'=c(TRUE, FALSE),
           'TRUE'=c(TRUE, FALSE),
           'FALSE'=c(FALSE, FALSE))

```

OR: |  

```{r, echo=FALSE}
data.frame('first_second'=c(TRUE, FALSE),
           'TRUE'=c(TRUE, TRUE),
           'FALSE'=c(TRUE, FALSE))

```

NOT: ! 
TRUE -> FALSE  
FALSE -> TRUE 

They are used in a following way: 

```{r, eval=FALSE}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)
firstLogical & secondLogical
firstLogical | secondLogical
! firstLogical

```

---

### Exercise: Subset by logical indexes
1. Return all the elements from your vector that are divisable by 3.  
2. Return all the elements from your vector that are divisable by 3 OR NOT divisable by 2.  
3. Return all the elements from your vector which are on EVEN positions (2,4,6,...)  



remember:  
     x == 0 - where is x equal to 0  
     x%%5 -> gives you the modulo while dividing x by 5  
     x != 0 - not equal to   

Solution?  

---
class: inverse, center, middle

## Functions 
#### Some useful functions  

---

### length - Gives you the length of the vector:    

```{r}
length(someothervector)

```

--

### unique -Gives you all unique elements in your vector:   
```{r}
unique(someothervector)
```

--
### table : gives you list of all elements and counts them  

```{r}
table(someothervector)
```

---
## Math:  


```{r}
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```


---

## more random math:  

### sample - gives you random numbers from a vector
```{r}
sample(1:100, 10)
```

--

### rnorm - gives you 10 random numbers from normal distribution with mean=0 and sd=1  

```{r}
rnorm(10, mean = 0, sd = 1)
```

--

analogously, rpois, runif

```{r,eval=FALSE}
runif(n = 10, min=10, max=20)
```

---
## Exercise!  

What will the following  produce? 

     x=seq(1,6, by=2) 
     x[x>4]           
     mean(x)          
     median(x)  
     x[c(T,F)]  
     rev(x)  


---
## Exercise!  
.pull-left[
Useful functions:  
     
     x=seq(1,6, by=1) 
     x[x>4]           
     mean(x)          
     median(x)  
     x[c(T,F)]  
     rev(x)  

]


.pull-right[
Create a sequence of numbers from 1 to 1000 and save it in a vector myseq.  
Part 1: What is the mean of the elements which are larger then the median?  
Part 2: What is the sum of every second element?  
Part 3: sum all the elements pairwise by summing the first one with the last one,
second one with second to last, and so on.

]



---
class: inverse, middle, center

## HELP!!??!!

---

### HELP!!??!!

```{r, eval=FALSE}
?runif
example(runif)
```
---
###  Exercise - estimate pi
Do the following: 

#### Create random points in 2D:  


1. Make a vector x with 1 000 000 random numbers uniformly distributed from -1 to 1.  
2. Make another vector y with 1 000 000 random numbers uniformly distributed from -1 to 1.  

Calculate percentage of points that fall within the circle:  
Calculate distance from center for each point!  

3. Create a vector dist that will calculate the result of x^2 + y^2   
4. Create a logical vector insideCircle that will check if dist is smaller then 1^2
5. Calculate the ratio of points that fall into the circle.  

     Useful functions:
     
     length(x) : vector length  
     as.numeric(x): conversion to num  
     mean(x): mean of all values in x
     


pi = 3.1415926535...  

