---
output: html_document
editor_options: 
  chunk_output_type: console
title: The Bayesian inference workshop
---
# Bayesian inference

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(reshape2)
library(magrittr)
library(readr)
library(stringr)
options(mc.cores = parallel::detectCores(), width = 140)
rstan_options(auto_write = TRUE)
```

## Clinical trial: classical approach

```{r}
(trial_outcomes <-
  tibble(condition = c("treatment","control"),
         cured = c(5,3),
         not_cured = c(5,8)))
```

### chi^2 test

Let's test whether the treatment effect is statistically significant.

```{r}
prop.test(as.matrix(trial_outcomes[2:3]))
```

### The GLM approach

We can also model this using a generalized linear model (GLM).

```{r}
glm_fit <- glm(cbind(cured,not_cured) ~ condition,
    family = binomial(), data=trial_outcomes)
summary(glm_fit)
```

## Clinical trial: Bayesian approach

### Analytical

Plot the posterior densities:

```{r}
trial_outcomes %>% group_by_all %>%
  do(x=seq(from=0,to=1,by=1e-3)) %>%
  unnest(x) %>%
  mutate(y=dbeta(x,cured+1,not_cured+1)) %>%
  ggplot() + geom_line(aes(x,y,color=condition))
```

Calculate the probability that the treatment works (increases the probability of getting cured).

Via integration:

```{r}
trial_outcomes %>%
  pivot_wider(names_from = condition, values_from=c(cured,not_cured)) %>%
  with(integrate(function(x)
    pbeta(x, cured_control+1, not_cured_control+1) *
    dbeta(x, cured_treatment+1, not_cured_treatment+1), lower=0, upper=1))
```

Via sampling:

```{r}
beta_samples <- trial_outcomes %>% group_by_all %>%
  do(i=seq(10000), p = rbeta(10000, .$cured+1, .$not_cured+1)) %>%
  unnest(i,p)
ggplot(beta_samples) + geom_density(aes(p,color=condition))
beta_samples_long <- beta_samples %>%
  pivot_wider(names_from = condition, values_from = p, id_cols = i) %>%
  mutate(treatment_works = treatment > control)
mean(beta_samples_long$treatment_works)
```

## MCMC

```{r}
mdl1 <- stan_model(model_code = "
  data {
    int outcomes[2,2];
  }
  parameters {
    vector<lower=0,upper=1>[2] p;
  }
  model {
    for (i in 1:2) {
      outcomes[i,1] ~ binomial(sum(outcomes[i]), p[i]);
    }
  }
  ")

mcmc_samples_1 <- sampling(mdl1, data = list(outcomes = as.matrix(trial_outcomes[2:3])), iter=10000, chains=10)
summary(mcmc_samples_1)
# Now repeat the graphs and the calculation
p_samples_1 <- rstan::extract(mcmc_samples_1)$p
colnames(p_samples_1) <- c("treatment", "control")
mean(p_samples_1[,"treatment"] > p_samples_1[,"control"])
```

### The probability of cure

#### Via filtering

```{r}
p_cured_samples_1 <-
  p_samples_1 %>% as_tibble %>%
    mutate(cure = treatment - control) %>%
    filter(cure > 0)
ggplot(p_cured_samples_1) + geom_density(aes(cure)) + lims(x=c(0,1))
mean(p_cured_samples_1$cure)
```

#### Via an amended model

Version 1:

```{r}
mdl2 <- stan_model(model_code = "
  data {
    int outcomes[2,2];
  }
  parameters {
    simplex[3] q;
  }
  transformed parameters {
    vector[2] p;
    p[1] = q[1]+q[2];
    p[2] = q[2];
  }
  model {
    for (i in 1:2) {
      outcomes[i,1] ~ binomial(sum(outcomes[i]), p[i]);
    }
  }
  ")

mcmc_samples_2 <- sampling(mdl2, data = list(outcomes = as.matrix(trial_outcomes[2:3])), iter=10000, chains=10)
q_samples_2 <- rstan::extract(mcmc_samples_2)$q
mean(q_samples_2[,1])
```

Version 2:

```{r}
mdl3 <- stan_model(model_code = "
  data {
    int outcomes[2,2];
  }
  parameters {
    vector<lower=0,upper=1>[2] q;
  }
  transformed parameters {
    vector[2] p;
    p[1] = q[1] * (1 - q[2]) + q[2];
    p[2] = q[2];
  }
  model {
    target += log(1-q[2]);
    for (i in 1:2) {
      outcomes[i,1] ~ binomial(sum(outcomes[i]), p[i]);
    }
  }
  ")

mcmc_samples_3 <- sampling(mdl3, data = list(outcomes = as.matrix(trial_outcomes[2:3])), iter=10000, chains=10)
p_samples_3 <- rstan::extract(mcmc_samples_3)$p
mean(p_samples_3[,1]-p_samples_3[,2])
```

### Priors

```{r}
mdl4 <- stan_model(model_code = "
  data {
    int outcomes[2,2];
  }
  parameters {
    simplex[3] q;
  }
  transformed parameters {
    vector[2] p;
    p[1] = q[1]+q[2];
    p[2] = q[2];
  }
  model {
    q ~ dirichlet([1, 2, 4]');
    for (i in 1:2) {
      outcomes[i,1] ~ binomial(sum(outcomes[i]), p[i]);
    }
  }
  ")

mcmc_samples_4 <- sampling(mdl4, data = list(outcomes = as.matrix(trial_outcomes[2:3])))
q_samples_4 <- rstan::extract(mcmc_samples_4)$q
mean(q_samples_4[,1])
```

## RNA-seq

Data loading

```{r}
expression_data <- read_tsv("GSE121757_SUPT4H1_HEK293_intron.tab.gz") %>%
  pivot_longer(matches("^(scramble|SUPT4H1)"), names_to = "sample", values_to = "read_count") %>%
  tidyr::extract(col=sample, into=c("condition", "replicate"), regex="(\\w+)_(\\d+)") %>%
  mutate(condition = ifelse(condition == "scramble", "control", "treatment")) %>%
  select(gene = ensembl_gene_id, condition, replicate, read_count) %>%
  group_by(condition, replicate) %>% mutate(library_size = sum(read_count)) %>% ungroup
ercc_expression <- expression_data %>% filter(str_starts(gene, "ERCC")) %>%
  rename(ercc_id = gene)
ercc_len <- read_tsv("ERCC92_len.tsv", col_names = c("ercc_id","length"))
ercc_mix_info <- read_tsv("ERCC_Controls_Analysis.txt") %>%
  select(ercc_id = `ERCC ID`, subgroup,
         control=`concentration in Mix 1 (attomoles/ul)`,
         treatment=`concentration in Mix 2 (attomoles/ul)`) %>%
  pivot_longer(cols = c(control,treatment), names_to = "condition", values_to = "concentration")
ercc_data <- ercc_expression %>%
  inner_join(ercc_len, by="ercc_id") %>%
  inner_join(ercc_mix_info, by=c("ercc_id","condition")) %>%
  mutate(sample_idx = group_indices(., condition, replicate),
         condition_idx = group_indices(., condition))
sample_info <- ercc_data %>% distinct(sample_idx, condition, condition_idx, replicate) %>% arrange(sample_idx)
```

Plotting

```{r}
ggplot(ercc_data) + geom_jitter(aes(concentration, read_count / library_size / length, color=condition)) +
  scale_x_log10() + scale_y_log10()

ggplot(ercc_data) + geom_jitter(aes(concentration, read_count / library_size, color=condition)) +
  scale_x_log10() + scale_y_log10()
```

```{r}
ercc_data %>% group_by(ercc_id, subgroup, condition) %>% summarise(rel_expr = mean(read_count/library_size)) %>% ungroup %>%
  pivot_wider(c(ercc_id, subgroup), names_from=condition, values_from=rel_expr) %>%
  ggplot() + geom_boxplot(aes(subgroup, log2(control/treatment)))
```

Rough estimation


```{r}
d <- ercc_data %>% filter(read_count > 0) %>%
  mutate(c = log(read_count / (library_size * concentration * length)))
ggplot(d) + geom_boxplot(aes(condition, c))
d %>% group_by(condition) %>% summarise(c=mean(c)) %>% pivot_wider(names_from=condition, values_from=c) %>%
  summarize(treatment-control)
```

Check the Poisson assumption

```{r}
ercc_mean_var <- ercc_data %>%
  mutate(expr = read_count / library_size) %>%
  group_by(ercc_id, condition) %>% summarize(mean = mean(expr), var=var(expr))
ggplot(ercc_mean_var) + geom_point(aes(mean,var, color=condition))
ggplot(ercc_mean_var) + geom_point(aes(mean,sqrt(var), color=condition))
```

Bayesian analysis

Poisson

```{r}
model_poisson <- stan_model(model_code = "
  data {
    // Number of spike-in measurements (combinations of sample × spike-in)
    int n_sp;
    // Number of samples
    int n_samples;
    // The expected concentration for each sample × spike-in
    vector[n_sp] concentration;
    // The observed read count for each sample × spike-in
    int read_count[n_sp];
    // Which sample the measurement comes from
    int<lower=1,upper=n_samples> sample[n_sp];
    // spike-in length
    int length[n_sp];
    // library size per measurement
    int library_size[n_sp];
  }
  
  parameters {
    // normalization constant per sample, proportional to the total RNA concentration
    vector<lower=0>[n_samples] rna_conc;
  }
  
  model {
    for (sp in 1:n_sp) {
      read_count[sp] ~ poisson(concentration[sp] * length[sp] * library_size[sp] / rna_conc[sample[sp]]);
    }
  }
")
```

```{r}
samples_poisson <- sampling(model_poisson, chains=4,
                     data = list(n_sp = nrow(ercc_data),
                                 n_samples = max(ercc_data$sample_idx),
                                 concentration = ercc_data$concentration,
                                 read_count = ercc_data$read_count,
                                 sample = ercc_data$sample_idx,
                                 length = ercc_data$length,
                                 library_size = ercc_data$library_size))

rstan::extract(samples_poisson)$rna_conc %>%
  melt(varnames = c("i","sample_idx"), value.name = "rna_conc") %>%
  inner_join(sample_info, by="sample_idx") %>%
  ggplot() + geom_density(aes(rna_conc)) + facet_grid(replicate ~ condition)
```


Negative binomial

```{r}
model_negbinom <- stan_model(model_code = "
  data {
    // Number of spike-in measurements (combinations of sample × spike-in)
    int n_sp;
    // Number of samples
    int n_samples;
    // The expected concentration for each sample × spike-in
    vector[n_sp] concentration;
    // The observed read count for each sample × spike-in
    int read_count[n_sp];
    // Which sample the measurement comes from
    int<lower=1,upper=n_samples> sample[n_sp];
    // spike-in length
    int length[n_sp];
    // library size per measurement
    int library_size[n_sp];
    // condition, per sample
    int <lower=1,upper=2> condition[n_samples];
    real phi;
  }
  
  parameters {
    // normalization constant per sample, proportional to the total RNA concentration
    vector<lower=0>[n_samples] rna_conc;
    // the overdispersion coefficient
    // vector<lower=0>[2] phi;
  }
  
  model {
    for (sp in 1:n_sp) {
      // read_count[sp] ~ poisson(concentration[sp] * length[sp] * library_size[sp] / rna_conc[sample[sp]]);
      read_count[sp] ~ neg_binomial_2(concentration[sp] * length[sp] * library_size[sp] / rna_conc[sample[sp]], phi);
      // read_count[sp] ~ neg_binomial_2(concentration[sp] * length[sp] * library_size[sp] / rna_conc[sample[sp]], phi[condition[sample[sp]]]);
    }
  }
")
```

```{r}
(samples_negbinom <- sampling(model_negbinom, chains=4,
                     data = list(n_sp = nrow(ercc_data),
                                 n_samples = max(ercc_data$sample_idx),
                                 concentration = ercc_data$concentration,
                                 read_count = ercc_data$read_count,
                                 sample = ercc_data$sample_idx,
                                 length = ercc_data$length,
                                 library_size = ercc_data$library_size,
                                 phi = 10,
                                 condition = sample_info$condition_idx)))

rstan::extract(samples_negbinom)$rna_conc %>%
  melt(varnames = c("i","sample_idx"), value.name = "rna_conc") %>%
  inner_join(sample_info, by="sample_idx") %>%
  ggplot() + geom_density(aes(rna_conc)) + facet_grid(replicate ~ condition)
```

Gamma for per-condition means

```{r}
model_gamma <- stan_model(model_code = "
  data {
    // Number of spike-in measurements (combinations of sample × spike-in)
    int n_sp;
    // Number of samples
    int n_samples;
    // The expected concentration for each sample × spike-in
    vector[n_sp] concentration;
    // The observed read count for each sample × spike-in
    int read_count[n_sp];
    // Which sample the measurement comes from
    int<lower=1,upper=n_samples> sample[n_sp];
    // spike-in length
    int length[n_sp];
    // library size per measurement
    int library_size[n_sp];
    // condition, per sample
    int <lower=1,upper=2> condition[n_samples];
  }
  
  parameters {
    // normalization constant per sample, proportional to the total RNA concentration
    vector<lower=0>[n_samples] rna_conc;
    // parameters of the gamma distribution describing rna_conc, per condition
    vector<lower=0>[2] gamma_alpha;
    vector<lower=0>[2] gamma_beta;
  }
  
  model {
    for (s in 1:n_samples) {
      rna_conc[s] ~ gamma(gamma_alpha[condition[s]], gamma_beta[condition[s]]);
    }
    for (sp in 1:n_sp) {
      read_count[sp] ~ poisson(concentration[sp] * length[sp] * library_size[sp] / rna_conc[sample[sp]]);
    }
  }
  
  generated quantities {
    real rna_conc_sample[2] = gamma_rng(gamma_alpha, gamma_beta);
    real ratio = rna_conc_sample[2] / rna_conc_sample[1];
  }
")
```

```{r}
samples_gamma <- sampling(model_gamma, chains=4, control = list(adapt_delta = 0.95),
                     data = list(n_sp = nrow(ercc_data),
                                 n_samples = max(ercc_data$sample_idx),
                                 concentration = ercc_data$concentration,
                                 read_count = ercc_data$read_count,
                                 sample = ercc_data$sample_idx,
                                 length = ercc_data$length,
                                 library_size = ercc_data$library_size,
                                 condition = sample_info$condition_idx))

        rstan::extract(samples_gamma)$rna_conc %>%
  melt(varnames = c("i","sample_idx"), value.name = "rna_conc") %>%
  inner_join(sample_info, by="sample_idx") %>%
  ggplot() + geom_density(aes(rna_conc)) + facet_grid(replicate ~ condition)
  
rstan::extract(samples_gamma)$ratio %>% tibble::enframe(name = NULL, value="ratio") %>%
  ggplot() + geom_density(aes(ratio))

rstan::extract(samples_gamma)$gamma_alpha %>%
  melt(varnames = c("i","condition"), value.name = "alpha") %>%
  ggplot() + geom_density(aes(alpha, color=as.factor(condition)))

rstan::extract(samples_gamma)$gamma_beta %>%
  melt(varnames = c("i","condition"), value.name = "beta") %>%
  ggplot() + geom_density(aes(beta, color=as.factor(condition)))
```