---
output: html_document
editor_options: 
  chunk_output_type: console
title: The Bayesian inference workshop
---
# Bayesian inference

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(rstan)
library(reshape2)
library(magrittr)
library(readr)
library(stringr)
options(mc.cores = parallel::detectCores(), width = 140)
rstan_options(auto_write = TRUE)
```

## Clinical trial: classical approach

```{r}
(trial_outcomes <-
  tibble(condition = c("treatment","control"),
         cured = c(5,3),
         not_cured = c(5,8)))
```

### Classical (frequentist) approach

Test whether the treatment effect is statistically significant.

### Analytical Bayesian approach

Plot the posterior densities `y=dbeta(x,cured+1,not_cured+1)`.

Calculate the probability that the treatment works (increases the probability of getting cured).

## MCMC

See <https://mc-stan.org/users/documentation/>

```{r}
mdl1 <- stan_model(model_code = "
  data {
    int outcomes[2,2];
  }
  parameters {
    vector<lower=0,upper=1>[2] p;
  }
  model {
    for (i in 1:2) {
      outcomes[i,1] ~ binomial(sum(outcomes[i]), p[i]);
    }
  }
  ")

mcmc_samples_1 <- sampling(mdl1, data = list(outcomes = as.matrix(trial_outcomes[2:3])))
p_samples_1 <- rstan::extract(mcmc_samples_1)$p
colnames(p_samples_1) <- c("treatment", "control")
```

Exercices:

* Plot the posterior distributions of p[1] and p[2]
* Calculate the probability that treatment works
* Calculate the probability that a patient won't improve without the treatment and will with the treatment
* Adjust the priors and see how that affects the result

## RNA-seq

Data loading

```{r}
expression_data <- read_tsv("GSE121757_SUPT4H1_HEK293_intron.tab.gz") %>%
  pivot_longer(matches("^(scramble|SUPT4H1)"), names_to = "sample", values_to = "read_count") %>%
  tidyr::extract(col=sample, into=c("condition", "replicate"), regex="(\\w+)_(\\d+)") %>%
  mutate(condition = ifelse(condition == "scramble", "control", "treatment")) %>%
  select(gene = ensembl_gene_id, condition, replicate, read_count) %>%
  group_by(condition, replicate) %>% mutate(library_size = sum(read_count)) %>% ungroup
ercc_expression <- expression_data %>% filter(str_starts(gene, "ERCC")) %>%
  rename(ercc_id = gene)
ercc_len <- read_tsv("ERCC92_len.tsv", col_names = c("ercc_id","length"))
ercc_mix_info <- read_tsv("ERCC_Controls_Analysis.txt") %>%
  select(ercc_id = `ERCC ID`, subgroup,
         control=`concentration in Mix 1 (attomoles/ul)`,
         treatment=`concentration in Mix 2 (attomoles/ul)`) %>%
  pivot_longer(cols = c(control,treatment), names_to = "condition", values_to = "concentration")
ercc_data <- ercc_expression %>%
  inner_join(ercc_len, by="ercc_id") %>%
  inner_join(ercc_mix_info, by=c("ercc_id","condition")) %>%
  mutate(sample_idx = group_indices(., condition, replicate),
         condition_idx = group_indices(., condition))
sample_info <- ercc_data %>% distinct(sample_idx, condition, condition_idx, replicate) %>% arrange(sample_idx)
```

* Inspect the data
* Construct a model

Bayesian model

```{r}
model_poisson <- stan_model(model_code = "
  data {
    // Number of spike-in measurements (combinations of sample × spike-in)
    int n_sp;
    // Number of samples
    int n_samples;
    // The expected concentration for each sample × spike-in
    vector[n_sp] concentration;
    // The observed read count for each sample × spike-in
    int read_count[n_sp];
    // Which sample the measurement comes from
    int<lower=1,upper=n_samples> sample[n_sp];
    // spike-in length
    int length[n_sp];
    // library size per measurement
    int library_size[n_sp];
  }
  
  parameters {
    // normalization constant per sample, proportional to the total RNA concentration
    vector<lower=0>[n_samples] rna_conc;
  }
  
  model {
    for (sp in 1:n_sp) {
      read_count[sp] ~ poisson(concentration[sp] * length[sp] * library_size[sp] / rna_conc[sample[sp]]);
    }
  }
")
```

```{r}
samples_poisson <- sampling(model_poisson, chains=4,
                     data = list(n_sp = nrow(ercc_data),
                                 n_samples = max(ercc_data$sample_idx),
                                 concentration = ercc_data$concentration,
                                 read_count = ercc_data$read_count,
                                 sample = ercc_data$sample_idx,
                                 length = ercc_data$length,
                                 library_size = ercc_data$library_size))

rstan::extract(samples_poisson)$rna_conc %>%
  melt(varnames = c("i","sample_idx"), value.name = "rna_conc") %>%
  inner_join(sample_info, by="sample_idx") %>%
  ggplot() + geom_density(aes(rna_conc)) + facet_grid(replicate ~ condition)
```

Exercises:

* Change the model to use the negative binomial distribution
* Calculate the average mean between the RNA concentrations